---
title: "Bioinfo Class8"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##K-means clustering example
testing kmeans() function
```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))

plot(x)
```
Use the kmeans() function setting k to 2 and nstart=20
```{r}
km <- kmeans(x, centers=2, nstart=20)
km
```

Inspect/print the results
Q. How many points are in each cluster?
```{r}
km$size
```

Q. What ‘component’ of your result object details
 - cluster size?
 - cluster assignment/membership?
 - cluster center?
```{r}
km$cluster

```
 cluster center
```{r}
km$centers

```
 
Plot x colored by the kmeans cluster assignment and
 add cluster centers as blue points
```{r}
plot(x, col=km$cluster, pch=16)

#points()
```
#hierarchical clustering
calculating point (dis)similarity as euclidean distance between observations
```{r}
dist_matrix <- dist(x)
```
```{r}
class(dist_matrix)
```

```{r}
dim(as.matrix(dist_matrix))
```

```{r}
hc <- hclust(d=dist_matrix)
```

```{r}
hc
```

```{r}
plot(hc)
```
"cut" to define clusters
```{r}
plot(hc)
abline(h=6, col="red")

```
```{r}
grps <- cutree(hc, h=6) # Cut by height h
table(grps)
```

```{r}
plot(x, col=grps)
```
```{r}
plot(x, col=cutree(hc, k=4))
```
# Step 1. Generate some example data for clustering
```{r}
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
```

# Step 2. Plot the data without clustering
```{r}
plot(x)

```
# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
```{r}
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
```
Your Turn!
Q. Use the dist(), hclust(), plot() and cutree()
 functions to return 2 and 3 clusters 
```{r}
dist_matrix <- dist(x)
dim(as.matrix(dist_matrix))
hc <- hclust(d=dist_matrix)
plot(hc)
abline(h=1.5, col="red")

```
```{r}
grps <- cutree(hc, h=1.5)
plot(x, col=grps)
```
for 2 groups
```{r}
dist_matrix <- dist(x)
dim(as.matrix(dist_matrix))
hc <- hclust(d=dist_matrix)
plot(hc)
abline(h=2.5, col="red")
```
```{r}
grps <- cutree(hc, h=2.5)
plot(x, col=grps)
```
Principal Component Analysis
```{r}
mydata <- read.csv("https://tinyurl.com/expression-CSV",
 row.names=1) 
head(mydata)
```
```{r}
pca <- prcomp(t(mydata), scale=TRUE) #PCAt9
attributes(pca)
```
##PC1 vs PC2 plot
```{r}
plot(pca$x[,1], pca$x[,2]) 
```
## Precent variance is often more informative to look at
```{r}
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```
```{r}
barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")
```
## A vector of colors for wt and ko samples
```{r}
#substr(x, start, stop) takes substrings from character vector
colvec <- as.factor( substr( colnames(mydata), 1, 2) ) 
colvec
```
```{r}
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 
text(pca$x[,1], pca$x[,2], labels=colnames(mydata))
```
#Worksheet
```{r}
x <- read.csv("UK_foods.csv")
dim(x)
```
```{r}
head(x)
```
##fix row names
```{r}
#can also do this with read.csv(row.names=1)
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```
```{r}
knitr::kable(x, caption="The full UK foods data table")
```
```{r}
par(mar=c(10, 4, 4, 2))
heatmap(as.matrix(x))
```
#PCA
```{r}
pca <- prcomp( t(x) )
summary(pca)
```

